Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2019-03-18 paper/paper.tex
added te Velthuis PB1 data *without* editing

\author{
  Alistair B. Russell,\afn{a}
  Jacob R. Kowalsky,\afn{a}
  Jesse D. Bloom\afn{a,b,c,\authfn{1}}
}

\affil{
  Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States\afn{a};
  Department of Genome Sciences, University of Washington, Seattle, United States\afn{b};
  Howard Hughes Medical Institute, Seattle, United States\afn{c}
}

@@

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.4\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.03\textwidth}
{\bf \Large B} \includegraphics[width=0.26\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
{\bf Reporter cells to identify and enrich infections that activate IFN expression.}
{\bf (A)} The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
{\bf (B)}
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

@@

\caption{
{\bf Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.}
{\bf (A)}
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
{\bf (B)}
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
{\bf (C)}
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
{\bf (D)}
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
{\bf (E)}
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

\caption{
{\bf Single-cell transcriptomics of IFN-enriched influenza-infected cells.}
{\bf (A)} 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
{\bf (B)} The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
{\bf (C)} Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
{\bf (D)} Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
{\bf (E)} Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
{\bf (F)} Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
{\bf (G)} Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.85\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
{\bf Viral genotypes and infection outcomes in single cells.}
Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1.
}
\label{fig:genotypes}

@@

%%% begin mutations figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[width=\linewidth]{figures/single_cell_figures/p_mutations.pdf}
}
\caption{
{\bf Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.}
{\bf (A)} 
Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
Cells infected by fully unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
{\bf (B)}
IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
{\bf (C)}
Specific viral defects associated with IFN induction.
The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
{\bf (D)}
There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
Throughout this figure, we only consider substitutions that are non-synonymous.
Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
}
\label{fig:mutations}

@@

\centerline{\includegraphics[width=0.9\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
\caption{
{\bf Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.}
{\bf (A)}
Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
The numbers indicate the median of four measurements for each viral mutant.
The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
{\bf (B)}
Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
}
\label{fig:validation}

@@

@@

We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (although some viral segments produce multiple splice forms, all mRNAs from a given segment share the same 5' end).
The sequences of these primers are in \SUPPFILE{primer_sequences}.
A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{boers2015micelle}.
We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume 100$\mu$l .
Both the common 3' primer and the multiplex mix of eight 5' primers were added to a final concentration of 0.5$\mu$M.
We performed 30 cycles of PCR, using an extension time of 2 minutes 15 seconds at 67$^{\circ}$C, and a melting temperature of 95$^{\circ}$C.
This melting temperature is lower than the standard 98$^{\circ}$C melting step suggested by the manufacturer for Kapa HiFi because we wanted to avoid collapse of emulsion integrity at high temperature.
We performed 30 cycles in order to saturate the reagents in each emulsion---unlike for open PCR reactions, the physical occlusion of amplified material in emulsion PCR reduces artifacts at high cycle numbers.

@@

@@

\section{Acknowledgments}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, AJ Velthuis, Katherine Xue, and Hannah Itell for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation (DRG-2227-15).
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.
The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.

@@

\begin{suppfig}
\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}
\caption{
{\bf Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.}
Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \cite{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\cite{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\cite{stetson2006type,honda2006type}.
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
\label{suppfig:mice}
\end{suppfig}

@@

\caption{
{\bf Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.}
{\bf (A)}
To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\cite{strahle2006sendai}.
The name of each of reporter cell line is indicated at the top of each row of plots.
At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
{\bf (B)}
The type I and type III IFN reporters are highly correlated in their activation.
An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
After 13 hours, cells were analyzed by flow cytometry.
Cells positive for either fluorescent reporter were further analyzed.
As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.
}
\label{suppfig:reportervalidation}
\end{suppfig}

>>>>

\author{
  Alistair B. Russell,\afn{a}
  Elizaveta Elshina,\afn{b}
  Jacob R. Kowalsky,\afn{a}
  Aartjan J. W. te Velthuis,\afn{b}
  Jesse D. Bloom\afn{a,c,d,\authfn{1}}
}

\affil{
  Basic Sciences Division and Computational Biology Program, Fred Hutchinson Cancer Research Center, Seattle, United States\afn{a};
  Division of Virology, Department of Pathology, Addenbrooke's Hospital, University of Cambridge, Cambridge, United Kingdom;
  Department of Genome Sciences, University of Washington, Seattle, United States\afn{c};
  Howard Hughes Medical Institute, Seattle, United States\afn{d}
}

@@

%%% start IFNrare figure %%%
\begin{figure}
\centerline{
{\bf \Large A}\includegraphics[width=0.4\textwidth,valign=t]{figures/IFN_stochastic/IFN_reporter/IFN_reporter.pdf}
\hspace{0.03\textwidth}
{\bf \Large B} \includegraphics[width=0.26\textwidth,valign=t]{figures/IFN_stochastic/Flow/ifn_percent.pdf}
}
\caption{
Reporter cells to identify and enrich infections that activate IFN expression.
(A) The reporter consists of an IFN promoter that drives expression of a cell-surface protein amenable to MACS and a fluorescent protein.
We created reporters with type I and type III IFN promoters (\SUPPFILE{reporter_sequences}).
In A549 cells, the reporters were efficiently activated by an IFN-inducing strain of Sendai virus (\SUPPFIG{reportervalidation}A).
(B)
Frequency of IFN induction upon infection with the influenza virus stock used in the single-cell studies in this paper, as quantified using the type III IFN reporter (see \SUPPFIG{IFNrare_flow} for full flow cytometry data).
The plot also shows uninfected cells, and cells infected with Sendai virus.
The limit of detection of 0.05\% is indicated with a dashed line, and numbers show the median of three measurements.
}
\label{fig:IFNrare}
\end{figure}
%%% end IFNrare figure %%%

@@

\caption{
Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
(A)
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
(B)
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
(C)
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
(D)
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
(E)
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
(A) 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
(B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
(C) Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
(D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
(F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
(G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.85\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
Viral genotypes and infection outcomes in single cells.
Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1.
}
\label{fig:genotypes}

@@

%%% begin mutations figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[width=\linewidth]{figures/single_cell_figures/p_mutations.pdf}
}
\caption{
Viral features associated with heterogeneity in infection outcome among cells for which we determined viral genotypes.
(A) 
Percent of all mRNA derived from virus, faceted by whether cells express unmutated copies of all eight genes.
Cells infected by fully unmutated virions exhibit less heterogeneity in viral burden as quantified by the Gini index (95\% confidence intervals are indicated).
(B)
IFN expression among cells expressing unmutated copies of all genes, and among cells with mutations or missing genes.
(C)
Specific viral defects associated with IFN induction.
The top panel show the percent of IFN- and IFN+ cells that fail to express each viral gene.
The middle and bottom panels show the percent of IFN- and IFN+ cells that have a deletion or amino-acid substitution in each gene, conditioned on the cell expressing that gene.
Numbers give P-values (Fisher's exact test) for rejecting the null hypothesis that percents are equal among IFN- and IFN+ cells. 
(D)
There is no association between IFN induction and the amount of viral mRNA in cells that express NS, but viral burden is associated with IFN induction among cells that lack NS.
Throughout this figure, we only consider substitutions that are non-synonymous.
Insertions are not shown as a mutation type as they are extremely rare (\FIG{genotypes}).
}
\label{fig:mutations}

@@

\centerline{\includegraphics[width=0.9\textwidth]{figures/Validation_Figure/ifn_plot.pdf}}
\caption{
Validation that IFN induction is increased by some of the mutations identified in the single-cell virus sequencing of IFN+ cells.
(A)
Percent of infected cells that become IFN+ after infection with a bulk stock of the indicated viral mutant, as determined using a reporter cell line.
The numbers indicate the median of four measurements for each viral mutant.
The limit of detection of 0.05\% is indicated with a dashed green line, and the median value for wild type is indicated with a dashed blue line.
Points are colored orange if the mutant virus stock induces IFN more frequently than the wild-type viral stock (one-sided t-test, $P < 0.01$), and blue otherwise.
(B)
Similar to the first panel, but validates the increased IFN induction for a large internal deletion in the PB1 gene.
See \SUPPFIG{validationflow} and \SUPPFIG{delqPCR} for more detailed experimental data.
The experiments in the two panels were performed on different days, and so numerical values can be reliably compared within panels but not between panels.
}
\label{fig:validation}

@@

\subsection{PB1 mutants}

%%% begin PB1 figure %%%
\begin{figure}

\centerline{\includegraphics[width=0.7\textwidth]{figures/PB1_figure.png}}
\caption{
Mutations D27N and T677A in the PB1 subunit affect polymerase activity in vitro. 
(A) Structural model of bat influenza A virus RNA polymerase (PDB 4WSB)~\cite{reich2014structural} superposed with the structure of the influenza B virus RNA polymerase (PDB 5MSG)~\cite{reich2017vitro}.
The locations of PB1 D27 and T677 (both red) of the influenza A virus polymerase relative to the 5' (blue) and 3' (orange) termini of the template and the PB1 active site (grey; PB1 act) are indicated.
The PA endonuclease (green; PA endo) and PB2 cap binding domain (pink; PB2 cap) are also indicated for reference. Part of the fingers subdomain of PB1 is not shown to reveal the template in the entry channel. 
(B) Influenza A virus RNA polymerase activity on a neuraminidase (NA) encoding vRNA template in HEK 293T cells. Steady state RNA levels were measured by primer extension, denaturing PAGE, and phosphorimaging. 
A PB1 active site control (PB1a) was used as negative control and background correction. 
The 5S rRNA signal was used as loading control. 
Other panels show western blot analysis of PB1, NP and GAPDH expression. 
Graph shows the average steady state RNA levels of three biological repeats. 
(C) Panel shows PAGE analysis of RT-PCR of short (17-200 nt long) influenza virus RNAs or mini viral RNAs (mvRNAs).
The length of the mini viral RNAs was extended by the adapters (adapt) present in the primers.
Graph shows the IFN-beta promoter activity in transfected HEK 293T cells as measured using a dual luciferase reporter assay.
(D) Influenza A virus polymerase activity on a segment 5-derived 246 nt long vRNA template. 
The top panel shows the steady state levels of vRNA template were analysed by primer extension and denaturing PAGE. 
The other two panels show the PB1 and tubulin expression levels analysed by western blot. 
The top graph shows the average steady state RNA levels of three biological repeats. 
The second graph shows the IFN-beta promoter activity in transfected HEK 293T cells as measured using a dual luciferase reporter assay. 
Error bars indicate standard deviation. 
P-values were determined by two-sided t-test.
}
\label{fig:PB1}

\end{figure}
%%% end PB1 figure %%%

Previous research has shown that aberrant RNA products, or mini viral RNAs (mvRNAs) {te Velthuis 2018}, produced during influenza A virus replication can trigger RIG-I activation and IFN production. To investigate if the PB1 mutations D27N and T667A could affect influenza A virus RNA polymerase activity we first identified their location and interactions with the template by superposing the pre-initiation complex of the bat influenza A virus RNA polymerase with the transcription initiation complex of the influenza B virus RNA polymerase (Fig. 7A). The model shows how the 3' terminus of the template enters the template entry channel above the PB1 active site and that T667A sits at tip of a helix that interacts with the 3' terminus as it enters the channel. By contrast, D27N resides deeper in the polymerase and close to the binding pocket of the 5' terminus. These observations suggest that both mutations may affect polymerase activity by changing the interactions of the RNA polymerase with the template.
To investigate how the two PB1 mutations affect polymerase activity on a full-length viral RNA template, we expressed the wildtype or mutant PB1 subunits along with PB2, PA, NP and the segment 6 vRNA in HEK 293T cells. Twenty-four hours after transfection, we extracted the total RNA and analysed the replication (vRNA level) and transcriptional (mRNA level) activity by primer extension. Interestingly, both mutations showed activities that were significantly different from wildtype (Fig. 7B), with D27N and T667A showing lower and higher viral replication and transcription, respectively. Both PB1 mutants showed higher activities than the PB1 active site control (PB1a) that we used to subtract background RNA levels. Moreover, both mutants were expressed to similar levels, as assessed by western blot. 
To investigate if the mutants produced any mvRNAs, we fractionated the total RNA to enrich for small RNA species (17-200 nt long) and performed an RT-PCR that amplified RNA species that contained both the 3' and 5' terminus of the template on equal amounts of RNA. This revealed that the D27N expressing cells contained higher levels of mvRNAs than the wildtype control and the T667A mutant (Fig. 7C). Analysis of the IFN levels showed that...
The low activity on the full-length segment 6 vRNA and higher mvRNA level produced on this template suggest that the D27N mutation in PB1 impaired polymerase processivity on full-length vRNAs. To confirm this, we replaced the segment 6 template with a segment 5-derived 246-nt long template (Fig. 7D). Under these conditions, the activity of the D27N mutant was similar to wildtype, while the activity of the T677A mutant proved again higher than wildtype, although not significantly higher in three biological repeats. Analysis of the IFN expression level in response to viral replication on the short template showed higher IFN induction by both PB1 mutants.

@@

We first performed a multiplex PCR reaction on 1 ng of the full-length 10X cDNA using a 3' primer complementary to the common 10X adaptor, and a multiplex mix of eight 5' primers, one specific for the mRNAs from each of the eight viral gene segments (although some viral segments produce multiple splice forms, all mRNAs from a given segment share the same 5' end).
The sequences of these primers are in \SUPPFILE{primer_sequences}.
A major concern during these PCRs is strand exchange (see \SUPPFIG{StrandExchange}) which would scramble the cell barcodes and mutations on viral cDNAs.
To reduce strand-exchange and hopefully obtain more even PCR amplification across segments, we performed emulsion PCRs using the Micellula DNA Emulsion Kit (Roboklon).
Emulsion PCR involves encapsidating DNA template molecules in a reverse-phase emulsion, with each template positive droplet serving as a microreactor.
This process physically separates disparate template molecules, preventing strand exchange and allowing each molecule to be amplified to exhaustion of its droplet's reagents without competing with the broader pool of PCR-amenable molecules~\cite{boers2015micelle}.
We performed the PCRs using Kapa HiFi Hotstart ReadyMix, supplementing the reactions with additional BSA to a final concentration of 0.1 mg/ml and using a volume of 100$\mu$l.
Both the common 3' primer and the multiplex mix of eight 5' primers were added to a final concentration of 0.5$\mu$M.
We performed 30 cycles of PCR, using an extension time of 2 minutes 15 seconds at 67$^{\circ}$C, and a melting temperature of 95$^{\circ}$C.
This melting temperature is lower than the standard 98$^{\circ}$C melting step suggested by the manufacturer for Kapa HiFi because we wanted to avoid collapse of emulsion integrity at high temperature.
We performed 30 cycles in order to saturate the reagents in each emulsion---unlike for open PCR reactions, the physical occlusion of amplified material in emulsion PCR reduces artifacts at high cycle numbers.

@@

\subsection{Structural analysis of PB1 mutants}
To locate the PB1 mutations in the influenza A virus RNA polymerase structure relative to the template and active site in \FIG{PB1}A, we superposed the bat influenza A virus RNA polymerase structure (PDB 4WSB)~\cite{reich2014structural}, which shows the 3' terminus of the template on the surface of the RNA polymerase, with the influenza B virus transcription initiation complex (PDB 5MSG)~\cite{reich2017vitro}, which shows the 3' terminus of the template in the template entry channel that leads towards the active site.
The structural alignment was performed in Pymol 1.8.7 using motifs A and C.

\subsection{Experimental analysis of PB1 mutants}
For the experimental analysis of the PB1 mutants in \FIG{PB1}, we used plasmids pcDNA-PB1, pcDNA-PA, pcDNA-PB2, and pcDNA-NP, which encode the WSN proteins that compose the polymerase complex~\cite{fodor2002single}; pPolI-NA, which encodes the viral RNA for the WSN NA~\cite{fodor2002single}; and pcDNA-PB1a, which encodes an inactive version of the WSN PB1 polymerase protein~\cite{vreede2004model}. 
To construct plasmids expressing mutant PB1 proteins D27N and T677A, the plasmid pcDNA-PB1 was subjected to site-directed mutagenesis. 
PB1 expression was analysed by western blot using antibody GTX125923 (GeneTex).

To analyse the activity of the PB1 mutants in cell culture, the plasmids expressing the WSN PA, PB2, NP and PB1 proteins were transfected into 293T cells together with the plasmid expressing the wildtype NA vRNA or a 246-nt long segment 5-based template~\cite{velthuis2018mini}.
Twenty-four hours post transfection, the RNA was extracted using Trizol (Invitrogen), and the steady state RNA levels assessed using reverse transcription with 32P-labelled oligonucleotides against the viral RNA species and ribosomal 5S RNA as described previously~\cite{velthuis2018mini, velthuis2016role}.
32P-derived signals were imaged using phosphorimaging on a Typhoon scanner and analysed using Prism (GraphPad).
In all experiments, the apparent RNA levels were background corrected using the PB1a mutant and normalized to the 5S rRNA loading control.

To measure the induction of the IFN-beta promoter during these RNP reconstitution assays, they were carried out in the presence of a plasmid expressing Renilla luciferase from a CMV promoter and a plasmid expressing Firefly luciferase from the IFN-beta promoter~\cite{velthuis2018mini}.
Twenty-four hours post transfection, cells were harvested, lysed and analysed using a DualGlo luciferase kit (Promega) according to the manufacturer’s instructions.
Samples were analysed using a GloMax (Promega).

@@

\section{Acknowledgments}
We thank Cole Trapnell, Jason Underwood, Robert Bradley, Daniel Stetson, Katherine Xue, and Hannah Itell for helpful suggestions.
We thank Andy Marty and the Fred Hutch Genomics Core for performing the deep sequencing.
This work was supported by the NIAID of the NIH under grant R01 AI127893 and a Burroughs Wellcome Fund Young Investigator in the Pathogenesis of Infectious Diseases grant to JDB.
ABR was supported by a postdoctoral fellowship from the Damon Runyon Cancer Research Foundation (DRG-2227-15).
JRK was supported by a Washington Research Foundation Undergraduate Research Fellowship and a Mary Gates undergraduate research scholarship from the University of Washington.
JDB is an Investigator of the Howard Hughes Medical Institute.
The funders had no role in study design, data collection and interpretation, or the decision to submit the work for publication.

@@

\begin{suppfig}
\includegraphics[width=\textwidth]{figures/IFN_stochastic/SteuermanReanalysis/plots/p_mRNA_counts.pdf}
\caption{
Re-analysis of a published single-cell transcriptomic study~\cite{steuerman2018dissection} of influenza-infected mice shows that only 5 of 1220 virus-infected cells express \emph{detectable} type I or type III IFN transcripts \textit{in vivo}.
Specifically, we downloaded the data from \cite{steuerman2018dissection} and identified influenza-infected cells using criteria similar to those described in \cite{steuerman2018dissection}.
Here we show statistics for the cells from the influenza-infected wild-type C57BL/6J mice, which were collected at 48 hours (two replicates) or 72 hours (one replicate)---we do not show cells from the control mice.
As shown in the left-most panel, the sequencing depth of \cite{steuerman2018dissection} was quite low, with only $\sim$1,500 mRNA counts per cell on average (this is 10 to 15-fold lower than the sequencing depth in the current study).
An important caveat is that this low sequencing depth could lead to a simple failure to detect IFN transcripts in some cells.
Nonetheless, there is detectable expression of key ISGs---which are induced by autocrine and paracrine signaling~\cite{stetson2006type,honda2006type}---in the majority of infected cells (the middle panel shows the total counts of IFIT1, ISG15, CCL5, and Mx1).
However, only 5 of 1220 cells express any detectable type I (IFN-$\alpha$ and IFN-$\beta$) or type III (IFN-$\lambda$) mRNAs, which are only induced by direct cellular detection of viral infection~\cite{stetson2006type,honda2006type}.
The full code that performs the re-analysis shown in this figure is at \url{https://github.com/jbloomlab/IFNsorted_flu_single_cell/tree/master/paper/figures/IFN_stochastic/SteuermanReanalysis/}.
}
\label{suppfig:mice}
\end{suppfig}

@@

\caption{
Validation of reporter cell lines (\FIG{IFNrare}A) to identify IFN+ cells.
(A)
To validate the IFN reporter cell lines, they were infected at high MOI with the Cantell strain of Sendai virus, which strongly activates IFN expression~\cite{strahle2006sendai}.
The name of each of reporter cell line is indicated at the top of each row of plots.
At 13 hours post-infection, activation of the IFN reporter was then monitored by flow cytometry using the marker indicated at the bottom of each plot (either a fluorescent protein or antibody staining for the cell-surface LNGFR$\Delta$C using a PE-conjugated anti-LNGFR antibody from Miltenyi Biotec).
Sendai infection efficiently activated the IFN reporter in all cases, with the strongest signal from the IFN-$\lambda$ reporter driving ZsGreen.
(B)
The type I and type III IFN reporters are highly correlated in their activation.
An A549 cell line was generated by transduction with both the IFN-$\beta$ and IFN-$\lambda$ reporters driving expression of mCherry and ZsGreen, respectively.
The cells were then infected with two different stocks of ``wild-type'' WSN influenza, or stocks with a deletion in PB1 or stop codons in NS1 (described later in the paper).
After 13 hours, cells were analyzed by flow cytometry.
Cells positive for either fluorescent reporter were further analyzed.
As shown in the FACS plots, expression of the \textit{IFNB1} and \textit{IFNL1} reporters is highly correlated in all cases.
}
\label{suppfig:reportervalidation}
\end{suppfig}
